%! TeX program = lualatex
\documentclass[../main.tex]{subfiles}

\usepackage{longtable}
\usepackage[table]{xcolor}

% Define colors for each month
\definecolor{Sep}{RGB}{226, 240, 217} % light green
\definecolor{Oct}{RGB}{222, 235, 247} % light blue
\definecolor{Nov}{RGB}{255, 242, 204} % light yellow
\definecolor{Dec}{RGB}{244, 219, 224} % light pink

\definecolor{Jan}{RGB}{226, 240, 217} % light green
\definecolor{Feb}{RGB}{222, 235, 247} % light blue
\definecolor{Mar}{RGB}{255, 242, 204} % light yellow
\definecolor{Apr}{RGB}{244, 219, 224} % light pink

\renewcommand{\arraystretch}{1.1}

% Define column types for right/top
\newcolumntype{T}{>{\centering\arraybackslash}p{1em}}
\newcolumntype{R}{>{\raggedleft\arraybackslash}p{2cm}}
\setlist[enumerate, 2]{label={\arabic*.}}

\newcommand{\classno}[1]{\parbox{2cm}{
\begin{tikzpicture}
  \draw[thick, dotted, gray] (0,0) circle[radius=2ex];
  \node at (0,0) {\color{main} #1};
\end{tikzpicture}
}}

\newcommand{\assessment}[1]{{\footnotesize #1 \faPencil*{}}}
\begin{document}

\section{Class schedules}

\begin{multicols}{2}
  \begin{enumerate}[wide, label={\textbf{Week \arabic*}.}]
    \item Fundamentals of modelling.
    \begin{enumerate}
      \item Dimension homogeneity.
      \item Scaling models.
      \item Allometry.
      \item Allometry continued.
    \end{enumerate}

  \item Recurrence.
    \begin{enumerate}[resume]
      \item Recurrence and their solutions.  Equilibria.
      \item Modelling with recurrence.
      \item Solving linear recursions.
      \item Problem-solving with recurrence.
    \end{enumerate}

  \item Conceptual understanding of DE.
    \begin{enumerate}[resume]
      \item Scientific computing session 1.
      \item DE. IVP and their solutions.
      \item \(y' = ky\) and qualitative information.
      \item Autonomous DE. Equilibra and stability.
    \end{enumerate}

  \item Working with DE.
    \begin{enumerate}[resume]
      \item The logistic model and its derivation.
      \item Modelling.
      \item Modelling continued.
      \item Classification. Linear first-order DE.
    \end{enumerate}

  \item DE continued.  Probability.
    \begin{enumerate}[resume]
      \item Linear first-order DE continued.
      \item Bernoulli substitution.
      \item \emph{Buffer}.
      \item Probability.
    \end{enumerate}

  \item Probability.
    \begin{enumerate}[resume]
      \item Probability continued.
      \item Conditional probability.
      \item Type-I and type-II errors.
      \item Independence.  Random variables.
    \end{enumerate}

  \item Reading week.

  \item Midterm week.
    \begin{enumerate}[resume]
      \item \emph{Buffer}.
      \item \emph{Buffer}.
      \item[{\faStar{}}] Midterm.
      \item[] No class.
    \end{enumerate}

  \item Random variables.
    \begin{enumerate}[resume]
      \item Discrete random variables.
      \item Discrete random variables continued.
      \item Continuous random variables.
      \item Continuous random variables continued.
    \end{enumerate}

  \item Simulation. Vectors and matricies.
    \begin{enumerate}[resume]
      \item Scientific computing session 2.
      \item Sampling and simulation.
      \item Vectors, linear systems and matricies.
      \item Modelling age-structured populations.
    \end{enumerate}

  \item Linearity. Eigen-theory.
    \begin{enumerate}[resume]
      \item Vector functions. Linearity.
      \item Linearity continued.
      \item Complex numbers.
      \item Eigenvalues and eigenvectors.
    \end{enumerate}

  \item Linear systems.
    \begin{enumerate}[resume]
      \item Eigenvalues and eigenvectors continued.
      \item Long-term behaviour of linear systems.
      \item Linear systems of recurrence.
      \item Linear systems of recurrence continued.
    \end{enumerate}

  \item Systems of DE.
    \begin{enumerate}[resume]
      \item Linear systems of DE.
      \item Linear systems of DE continued.
      \item Lokta-Volterra.
    \end{enumerate}

  \item Review.
  \end{enumerate}
\end{multicols}
\clearpage

\section*{Detailed Learning Outcomes}
\begin{itemize}
  \item \textbf{Dimensional homogeneity}.  Students should be able to
    \begin{enumerate}
      \item identify whether a given equation is dimensionally homogeneous.
      \item deduce dimensions (units) of a unknown quantity from an given dimensionally homogeneous equation.
    \end{enumerate}

  \item \textbf{Scaling models} \(y = c x^{\alpha}\) for \(c \ne 0\) or, equivalently, \(y \propto x^{\alpha}\). Students should be able to 
    \begin{enumerate}
      \item apply log-log transformation to obtain \(\ln(y) = \alpha \ln(x) + \ln(c)\).
      \item transform a power rule model in log-log form \(\ln(y) = \alpha \ln(x) + \ln(c)\) back into a power rule expression \(y = c x^{\alpha}\).
      \item plot given power rule models in log-log scale.
      \item identify power rule models from their log-log plots.
      \item use algebraic manipulations to relates two or more quantities using power rule models.
    \end{enumerate}

  \item \textbf{Scaling relations}, i.e., isometric vs allometric scaling.  Students should be able to 
    \begin{enumerate}
      \item identify if given power rule models exhibit isometric or allometric scaling relation.
      \item calculate proportions of increase/decrease given one or more related power rule models.
      \item use a \textbf{combination of skills} from dimension homogeneity, power rule models and scaling relations to solve problems.
    \end{enumerate}

  \item \textbf{Linear recursions} of the form \(u_{t} = \lambda u_{t-1} + b\) with and without initial conditions.  Students should be able to
    \begin{enumerate}
      \item manually calculate the first few terms of a linear recursion.
      \item model using linear recursions.
      \item interpret models involving linear recursions.
      \item solve such linear recursions with or without an initial condition.
      \item find unknown parameters in given linear recursions satisfying given conditions.
      \item long-term and qualitative behaviours of linear recursions by studying its solution using calculus (restricted to Calc~1000A techniques).
    \end{enumerate}

  \item \textbf{Differential equations} of a single unknown function. Students should be able to 
    \begin{enumerate}
      \item verify if a proposed function is a solution to an \textbf{initial-value problem}. Such problems are not necessarily restricted to linear first-order differential equations.
      \item use calculus to investigate qualitative behaviour (increasing/decreasing) of differential equations.
    \end{enumerate}

  \item \textbf{Modelling with differential equations}. 
  \item use dimension homogeneity to determine if given differential equations is valid or not.
    \begin{enumerate}
      \item Autonomous equations.  Students should be able to
        \begin{enumerate}
          \item recognize autonomous equations.
          \item find equilibria of autonomous equations algebraically or graphically.
          \item use phase-line plots to determine stability of equilibria.
          \item understand the logistic model and its assumptions.
        \end{enumerate}
      \item Conservation law \(\frac{d}{dt} (\text{quantity}) = (\text{net rate of production}) - \text{net rate of loss})\). Students should be able to
        \begin{enumerate}
          \item model real-life phenomenons using differential equations.
          \item use the conservation law to model mixing problems in a single variable.
          \item interpret models of differential equations using the conservation law.
          \item use differential equations to model a metabolism problem. See Geoff's notes Example~2.6 on page 19.
        \end{enumerate}
    \end{enumerate}

  \item Solving \textbf{differential equations}. Students should be able to
    \begin{enumerate}
      \item recognize and classify differential equations.
      \item identify an appropriate integrating factor given linear first-order differential equations.
      \item solve linear first-order differential equations by \emph{showing the entire process} of undoing the product rule.
      \item solve suitable non-linear first-order differential equation using \textbf{Bernoulli substitution}.
      \item find general or specific solutions depending on the context. 
    \end{enumerate}

  \item Fundamentals of \textbf{Probability}.  Students should be able to 
    \begin{enumerate}
      \item understand sample space and the definition of probability of events.
      \item describe elements of the sample space of a probability.
      \item calculate probability using basic counting.
      \item calculate probability from Venn diagrams.
      \item apply fundamental properties of probabilities using set relations.
      \item understand \textbf{mutually exclusive and exhaustive events} and their algebraic relations.
      \item understand \textbf{conditional probabilities}, \textbf{independence} and the \textbf{law of total probability} and their algebraic relations.
    \end{enumerate}

  \item \textbf{Conditional probabilities}. Students should be able to
    \begin{enumerate}
      \item calculate conditional probabilities using the definition \(\mathbb{P}(A \mid B) = \frac{\mathbb{P}(A \cap B)}{\mathbb{\mathbb{B}}}\).
      \item calculate conditional probabilities using Venn diagrams.
      \item understand and calculate probabilities of type-I and type-II errors and their related probabilities.
      \item apply conditional probabilities to real-life applications.
    \end{enumerate}

  \item \textbf{Random variables}. Students should be able to 
    \begin{enumerate}
      \item understand the definition random variables and associated terminologies and notations.
      \item list elements of the sample space of random variables.
      \item express random variables as functions.
      \item distinguish discrete versus continuous random variable.
      \item understand \textbf{probability mass functions} and \textbf{cumulative distribution functions} of \emph{discrete} random variables.
      \item understand \textbf{probability density functions} and \textbf{cumulative distribution functions} of \emph{continuous} random variables.
      \item calculate and/or sketch probability mass/density functions from cumulative distribution functions using calculus (restricted to Calc~1000A techniques).
      \item understand the relation between probability mass/density functions and their corresponding cumulative distribution functions.
      \item determine the validity of proposed probability mass/density functions and 
      \item calculate \textbf{mean}, \textbf{variance} and \textbf{standard deviation} from random variables (restricted to Calc~1000A techniques for continuous random variables).
       \item understand uniform distribution \(U(a,b)\).
       \item determine unknown parameter of \(U(a,b)\) satisfying certain given requirements.
    \end{enumerate}
    
  \item \textbf{Working with (sample) data}. Students should be able to
     \begin{enumerate}
       \item calculate sample mean, sample variance and sample standard deviation.
       \item calculate \(z\)-score of sample data.
       \item use \(z\)-scores to compare inherently random data.
       \item understand the purpose of each component of a statistical simulation.
       \item understand how to choose parameters for a statistical simulation.
       \item use Python to perform statistical simulations.
     \end{enumerate}

   \item \textbf{Vectors} and \textbf{systems of linear equations}. Students should be able to
     \begin{enumerate}
       \item model multi-variate quantities using vectors.  In particular, be able to distinguish state vectors and change vectors from given context.
       \item express linear equations as matrix-vector equations, and vice versa.
       \item perform linear operations (addition, subtractions and scalar multiplication) with vectors algebraically and pictorially.
       \item perform matrix operations (addition, subtractions, multiplication and exponentiation) with vectors.
       \item perform matrix-vector operations operations.
       \item understand functions whose input and output are vectors.
       \item manually calculate outputs of functions whose inputs are vectors.
     \end{enumerate}

   \item Applications of \textbf{Linear systems of recurrence} of the form \(\vec{v}(t) = A \vec{v}(t+1)\).
     \begin{enumerate}
       \item understand recurrences of the form \(\vec{v}(t) = A \vec{v}(t+1)\) with and without initial conditions.
       \item understand the vector field of matricies \(A\) by associating each vector \(\vec{v} \in \mathbb{R}^{2}\) with \((A - I) \vec{v}\), typically scaled.
       \item use vector fields to visually identify solutions to linear systems of recurrence.
       \item model \textbf{age-structured populations and similar phenomena} using linear system of recurrence and understand their transition diagrams.
       \item draw the transition diagram of given age-structure linear systems of recurrence.
     \end{enumerate}

   \item \textbf{Eigenvalues and eigenvectors}. Students should be able to
     \begin{enumerate}
       \item understand complex numbers for the purpose of finding eigenvalues of \(2 \times 2\) systems.
       \item understand, algebraically and pictorially, the effect of matricies acting on its eigenvectors for both real and complex eigenvectors.
       \item write down the determinant of \(2 \times 2\) and \(3 \times 3\) matricies.
       \item write down and solve the \textbf{characteristic polynomial} of \(2 \times 2\) and \(3 \times 3\) systems manually.
       \item find all eigenvalues (real and/or complex) and their associated eigenvectors of linear systems.
       \item determine asymptotic behaviour of linear recurrence \(\vec{v}(t+1) = A \vec{v}(t)\) by identifying dominant eigenvalues and eigenvectors.
       \item solve recurrence \(\vec{v}(t+1) = A \vec{v}(t)\) with given initial condition.
     \end{enumerate}

   \item \textbf{Systems of differential equations}. Students should be able to 
     \begin{enumerate}
       \item solve IVP of homogeneous \(2 \times 2\) systems of DEs with distinct real eigenvalues.
       \item understand the Lokta-Volterra model.
     \end{enumerate}
\end{itemize}
\clearpage

\begin{longtable}[t]{|T|R|R|R|R|R|R|R|}
\multicolumn{8}{c}{\textbf{2026 Winter}, \thiscourse, Mo/We/Th/Fr} \\
\toprule
W    & Sun    & \textbf{Mon}             & Tue                     & \textbf{Wed}          & \textbf{Thu}        & \textbf{Fri}            & Sat    \\
\midrule
\endfirsthead
\toprule
W    & Sun    & \textbf{Mon}             & Tue                     & \textbf{Wed}          & \textbf{Thu}        & \textbf{Fri}            & Sat    \\
\midrule
\endhead
\midrule
\endfoot
\bottomrule
\endlastfoot

\textbf{1 } & Jan 4  & \cellcolor{Jan}5        & 6                      & \cellcolor{Jan}7      & \cellcolor{Jan}8    & \cellcolor{Jan}9        & 10     \\
            &        & \cellcolor{Jan}\classno{1} &                     & \cellcolor{Jan}\classno{2} & \cellcolor{Jan}\classno{3} & \cellcolor{Jan}\classno{4} &        \\\midrule
\textbf{2 } & 11     & \cellcolor{Jan}12       & 13                     & \cellcolor{Jan}14     & \cellcolor{Jan}15   & \cellcolor{Jan}16       & 17     \\
            &        & \cellcolor{Jan}\classno{5} &                     & \cellcolor{Jan}\classno{6} & \cellcolor{Jan}\classno{7} & \cellcolor{Jan}\classno{8} &        \\\midrule
\textbf{3 } & 18     & \cellcolor{Jan}19       & 20                     & \cellcolor{Jan}21     & \cellcolor{Jan}22   & \cellcolor{Jan}23       & 24     \\
            &        & \cellcolor{Jan}\classno{9} &  \assessment{Quiz 1}                    & \cellcolor{Jan}\classno{10} & \cellcolor{Jan}\classno{11} & \cellcolor{Jan}\classno{12} &        \\\midrule
\textbf{4 } & 25     & \cellcolor{Jan}26       & 27                   & \cellcolor{Jan}28     & \cellcolor{Jan}29   & \cellcolor{Jan}30       & 31     \\
            &        & \cellcolor{Jan}\classno{13} &              & \cellcolor{Jan}\classno{14} & \cellcolor{Jan}\classno{15} & \cellcolor{Jan}\classno{16} &        \\\midrule
\textbf{5 } & Feb 1  & \cellcolor{Feb}2       & 3                     & \cellcolor{Feb}4      & \cellcolor{Feb}5    & \cellcolor{Feb}6        & 7      \\
            &        & \cellcolor{Feb}\classno{17} & \assessment{Quiz 2}                   & \cellcolor{Feb}\classno{18} & \cellcolor{Feb}\classno{19} & \cellcolor{Feb}\classno{20} &        \\\midrule
\textbf{6 } & 8      & \cellcolor{Feb}9        & 10                     & \cellcolor{Feb}11     & \cellcolor{Feb}12   & \cellcolor{Feb}13       & 14     \\
            &        & \cellcolor{Feb}\classno{21} & \assessment{Quiz 3}                  & \cellcolor{Feb}\classno{22} & \cellcolor{Feb}\classno{23} & \cellcolor{Feb}\classno{24} &        \\\midrule
\textbf{7 } & 15     & 16                      & 17                     & 18                    & 19                  & 20                      & 21     \\
            &        & \multicolumn{5}{c|}{Reading Week} &                                                                                               \\\midrule
\textbf{8 } & 22     & \cellcolor{Feb}23       & 24                     & \cellcolor{Feb}25     & \cellcolor{Feb}26   & 27       & 28     \\
            &        & \cellcolor{Feb}\classno{25} & \assessment{Quiz 4}                   & \cellcolor{Feb}\classno{26} & \cellcolor{Feb}\parbox{2cm}{Midterm\faStar} & \textit{No class} &        \\\midrule
\textbf{9 } & Mar 1  & \cellcolor{Mar}2        & 3                      & \cellcolor{Mar}4      & \cellcolor{Mar}5    & \cellcolor{Mar}6        & 7      \\
            &        & \cellcolor{Mar}\classno{27} &                    & \cellcolor{Mar}\classno{28} & \cellcolor{Mar}\classno{29} & \cellcolor{Mar}\classno{30} &        \\\midrule
\textbf{10} & 8      & \cellcolor{Mar}9        & 10                     & \cellcolor{Mar}11     & \cellcolor{Mar}12   & \cellcolor{Mar}13       & 14     \\
            &        & \cellcolor{Mar}\classno{31} &                    & \cellcolor{Mar}\classno{32} & \cellcolor{Mar}\classno{33} & \cellcolor{Mar}\classno{34} &        \\\midrule
\textbf{11} & 15     & \cellcolor{Mar}16       & 17                     & \cellcolor{Mar}18     & \cellcolor{Mar}19   & \cellcolor{Mar}20       & 21     \\
            &        & \cellcolor{Mar}\classno{35} & \assessment{Quiz 5}                  & \cellcolor{Mar}\classno{36} & \cellcolor{Mar}\classno{37} & \cellcolor{Mar}\classno{38} &        \\\midrule
\textbf{12} & 22     & \cellcolor{Mar}23       & 24                     & \cellcolor{Mar}25     & \cellcolor{Mar}26   & \cellcolor{Mar}27       & 28     \\
            &        & \cellcolor{Mar}\classno{39} & \assessment{Quiz 6} & \cellcolor{Mar}\classno{40} & \cellcolor{Mar}\classno{41} & \cellcolor{Mar}\classno{42} &        \\\midrule
\textbf{13} & 29     & \cellcolor{Mar}30       & 31                     & \cellcolor{Apr}1      & \cellcolor{Apr}2    & 3                   & 4      \\
            &        & \cellcolor{Mar}\classno{43} & \assessment{Quiz 7}                  & \cellcolor{Apr}\classno{44} & \cellcolor{Apr}\classno{45} &      &        \\\midrule
\textbf{14} & 5      & \cellcolor{Apr}6        & 7                      & \cellcolor{Apr}8      & \cellcolor{Apr}9    & 10       & 11     \\
            &        & \cellcolor{Apr}\classno{46} & \assessment{Quiz 8}                 & \cellcolor{Apr}\classno{47} & \cellcolor{Apr}\classno{48} & &        \\
\end{longtable}
\clearpage

\begin{longtable}[t]{|T|R|R|R|R|R|R|R|}
\multicolumn{8}{c}{\textbf{2025 Winter}, \thiscourse, Mo/We/Th/Fr} \\
\toprule
W    & Sun    & \textbf{Mon}             & Tue                     & \textbf{Wed}          & \textbf{Thu}        & \textbf{Fri}            & Sat    \\
\midrule
\endfirsthead
\toprule
W    & Sun    & \textbf{Mon}             & Tue                     & \textbf{Wed}          & \textbf{Thu}        & \textbf{Fri}            & Sat    \\
\midrule
\endhead
\midrule
\endfoot
\bottomrule
\endlastfoot

\textbf{1 } & Jan 5  & \cellcolor{Jan}6        & 7                      & \cellcolor{Jan}8      & \cellcolor{Jan}9    & \cellcolor{Jan}10       & 11     \\
            &        & \cellcolor{Jan}\classno{1} &                     & \cellcolor{Jan}\classno{2} & \cellcolor{Jan}\classno{3} & \cellcolor{Jan}\classno{4} &        \\\midrule
\textbf{2 } & 12     & \cellcolor{Jan}13       & 14                     & \cellcolor{Jan}15     & \cellcolor{Jan}16   & \cellcolor{Jan}17       & 18     \\
            &        & \cellcolor{Jan}\classno{5} &                     & \cellcolor{Jan}\classno{6} & \cellcolor{Jan}\classno{7} & \cellcolor{Jan}\classno{8} &        \\\midrule
\textbf{3 } & 19     & \cellcolor{Jan}20       & 21                     & \cellcolor{Jan}22     & \cellcolor{Jan}23   & \cellcolor{Jan}24       & 25     \\
            &        & \cellcolor{Jan}\classno{9} &                     & \cellcolor{Jan}\classno{10} & \cellcolor{Jan}\classno{11} & \cellcolor{Jan}\classno{12} &        \\\midrule
\textbf{4 } & 26     & \cellcolor{Jan}27       & 28                     & \cellcolor{Jan}29     & \cellcolor{Jan}30   & \cellcolor{Jan}31       & Feb 1  \\
            &        & \cellcolor{Jan}\classno{13} &                    & \cellcolor{Jan}\classno{14} & \cellcolor{Jan}\classno{15} & \cellcolor{Jan}\classno{16} &        \\\midrule
\textbf{5 } & 2      & \cellcolor{Feb}3        & 4                      & \cellcolor{Feb}5      & \cellcolor{Feb}6    & \cellcolor{Feb}7        & 8      \\
            &        & \cellcolor{Feb}\classno{17} &                    & \cellcolor{Feb}\classno{18} & \cellcolor{Feb}\classno{19} & \cellcolor{Feb}\classno{20} &        \\\midrule
\textbf{6 } & 9      & \cellcolor{Feb}10       & 11                     & \cellcolor{Feb}12     & \cellcolor{Feb}13   & \cellcolor{Feb}14       & 15     \\
            &        & \cellcolor{Feb}\classno{21} &                    & \cellcolor{Feb}\classno{22} & \cellcolor{Feb}\classno{23} & \cellcolor{Feb}\classno{24} &        \\\midrule
\textbf{7 } & 16     & 17                      & 18                     & 19                    & 20                  & 21                      & 22     \\
            &        & \multicolumn{5}{c|}{Reading Week} &                                                                                               \\\midrule
\textbf{8 } & 23     & \cellcolor{Feb}24       & 25                     & \cellcolor{Feb}26     & \cellcolor{Feb}27   & \cellcolor{Feb}28       & Mar 1  \\
            &        & \cellcolor{Feb}\classno{25} &                    & \cellcolor{Feb}\classno{26} & \cellcolor{Feb}\parbox{2cm}{Midterm\faStar} & \cellcolor{Feb}\classno{27} &        \\\midrule
\textbf{9 } & 2      & \cellcolor{Mar}3        & 4                      & \cellcolor{Mar}5      & \cellcolor{Mar}6    & \cellcolor{Mar}7        & 8      \\
            &        & \cellcolor{Mar}\classno{28} &                    & \cellcolor{Mar}\classno{29} & \cellcolor{Mar}\classno{30} & \cellcolor{Mar}\classno{31} &        \\\midrule
\textbf{10} & 9      & \cellcolor{Mar}10       & 11                     & \cellcolor{Mar}12     & \cellcolor{Mar}13   & \cellcolor{Mar}14       & 15     \\
            &        & \cellcolor{Mar}\classno{32} &                    & \cellcolor{Mar}\classno{33} & \cellcolor{Mar}\classno{34} & \cellcolor{Mar}\classno{35} &        \\\midrule
\textbf{11} & 16     & \cellcolor{Mar}17       & 18                     & \cellcolor{Mar}19     & \cellcolor{Mar}20   & \cellcolor{Mar}21       & 22     \\
            &        & \cellcolor{Mar}\classno{36} &                    & \cellcolor{Mar}\classno{37} & \cellcolor{Mar}\classno{38} & \cellcolor{Mar}\classno{39} &        \\\midrule
\textbf{12} & 23     & \cellcolor{Mar}24       & 25                     & \cellcolor{Mar}26     & \cellcolor{Mar}27   & \cellcolor{Mar}28       & 29     \\
            &        & \cellcolor{Mar}\classno{40} &                    & \cellcolor{Mar}\classno{41} & \cellcolor{Mar}\classno{42} & \cellcolor{Mar}\classno{43} &        \\\midrule
\textbf{13} & 30     & \cellcolor{Mar}31       & Apr 1                  & \cellcolor{Apr}2      & \cellcolor{Apr}3    & \cellcolor{Apr}4        & 5      \\
            &        & \cellcolor{Mar}\classno{44} &                    & \cellcolor{Apr}\classno{45} & \cellcolor{Apr}\classno{46} & \cellcolor{Apr}\classno{47} &        \\
\end{longtable}
\clearpage

\end{document}
