%! TeX program = lualatex
\documentclass[../main.tex]{subfiles}
\begin{document} \section{Complex numbers}

Complex numbers come from the theoretical need to solve equations such as \(x^{2} = -1\) but are fundamentally useful in all sciences.  In this section, we learn complex numbers for the purpose of understanding how vectors move in a vector field by calculating eigenvalues in the next section.

\begin{definition}[complex numbers]
  The symbol \(i\) is called \hlmain{\emph{the} imaginary number}, and we define \hlwarn{Edited: ``and define'' \(\to\) ``and we define''}
  % is forced to satisfy
  \[
    \sqrt{-1} = \underline{\hspace{1cm}}
    \quad\text{and}\quad 
    i^{2} = \underline{\hspace{1cm}}.
  \]

  \hlmain{Complex numbers} are \(a + bi\) where \(a,b\) are real numbers. In this form, \(a\) is called its real part, and \(b\) its imaginary part.
\end{definition}

\faStar{} We use complex numbers and the quadratic formula to find roots of characteristic polynomials of matrices.
\begin{example}
  Let \(\lambda\) be an unknown. Solve \((\lambda - 5) (\lambda^{2} + 5) = 0\).

  \blanks{10}
\end{example}

\begin{example}
  Let \(\lambda\) be an unknown. Solve \(\lambda^{2} + \lambda + 3 = 0\).

  \blanks{20}
\end{example}
\clearpage

We will need to do a little of arithmetic with complex numbers down the road.

\begin{definition}
  Arithmetic rules of complex numbers.
  \begin{align*}
    i^{2} &= -1 \\
    (a + bi) + (c + di) &= (a + c) + (b + d)i \\
    (a + bi) - (c + di) &= (a - c) + (b - d)i \\
    (a + bi)(c + di) &= (ac - bd) + (ad + bc)i.
  \end{align*}
\end{definition}

To simplify a complex number is to write it in the form of \(a + bi\).

\begin{example}
  Perform the following operations and simplify.

  \begin{itemize}[wide]
    \item \(\left(\frac{1}{3} + i\right)\left(1 - \frac{i}{2}\right) = \)

      \blanks{10}
      
    \item \((3 + 4i)(3 - 4i) = \)

      \blanks{10}

  \end{itemize}
\end{example}

\begin{definition}
  Complex division \(\frac{a + bi}{c + di} = \frac{(a + bi)(c-di)}{c^{2} + d^{2}}\).
\end{definition}

\begin{example}
  Simplify \(\frac{1 - 2i}{3 + 4i}\).

  \blanks{5}
\end{example}

\end{document}
