%! TeX program = lualatex
\documentclass[../main.tex]{subfiles}
\begin{document} \section{The logistic model}

We develop and analyze a more realistic population model called the \hlmain{logistic equation}.  It is
\begin{itemize}
  \item sophisticated enough to account for \emph{crowding}, 
  \item simple enough that it can be modified to account for additional factors, and
  \item one of two fundamental models in biology. The other one is called the Lotka-Volterra model.
\end{itemize}

{\footnotesize \faYoutube{} Not too excited about population? A solution to the logistic equation, called the Sigmoid curve, is used in early theory of machine learning as an activation function for neural networks. Here is a 20-minute video on neural networks \hlwarn{Edited: ``neural network'' \(\to\) ``neural networks''; moved the trailing period inside the footnote-size group} by Grant Sanderson (3Blue1Brown): \url{https://youtu.be/aircAruvnKk}.}

Everyday scientific developments are results of attempting to extend an existing idea just a tiny bit further.  Thinking back to the very first lecture of the term, we discussed how \emph{assumptions and simplifications} affect the validity and utility of models.  To advance science, we attempt to tweak assumptions to capture factors simplified away by existing ideas. 
% when assumptions break, our model is no longer valid. can't put diesel into a gas car.
% when there are too many simplifications, our model is no longer useful. 

In Example~\ref{ex:diff-eq-intro}, we discussed the population model \(N'(t) = (B - D) N(t)\) with the assumption that \emph{per capita birth and death rates} were \underline{\hspace{2in}}.  This assumption makes the math simple but not terribly realistic.  We demonstrate how a crowding constraint can be incorporated into a model by retracing the development of the logistic model.  The following is taken from \emph{How Ecologists Develop the Logistic Model} by Pfaff, PRIMUS (\url{https://doi.org/10.1080/10511970.2024.2310249}).

\blanklines{30}
\clearpage

\begin{definition}[logistic model]
  The population model \(N(t)\) with a growth factor \(r > 0\) and a carrying capacity \(K\) satisfies the differential equation
  \begin{equation} \label{eq:diff-eq-logistic}
    N'(t) = r N(t) \left( 1 - \frac{N(t)}{K} \right).
  \end{equation}

  The carrying capacity \(K\) is a constant representing the size of a population that the modelled environment can sustain over time.
\end{definition}

Visualization. \faExternalLink*{} \url{https://colab.research.google.com/drive/193VoMFnAjlcJ7laTlG4G2ioIw-nCkZ6j}

We advertised that the logistic model accounts for the crowding effect. Is it really true? We analyze the logistic model \emph{without solving the differential equation}. 

\begin{example} \label{ex:diff-eq-logistic-fact-check}
  The qualitative analysis we learned last week can pick out the crowding effect in the logistic model.

  \begin{enumerate}[wide]
    \item Suppose \(N(\text{some time})\) is below and far away from \(K\). How does the population change \emph{according} to the model?

      \blanklines{10}

    \item Suppose \(N(\text{some time})\) is below but close to \(K\). How does the population change \emph{according} to the model?

      \blanklines{10}

    \item Suppose \(N(\text{some time})\) is larger than \(K\). How does the population change \emph{according} to the model?

      \blanklines{10}
  \end{enumerate}
\end{example}
\clearpage

The logistic model is relatively simple and can be easily modified to account for additional factors.

How would you modify the logistic equation to account for
\begin{enumerate}[wide]
  \item variable growth rate and variable carrying capacity, potentially due to environmental changes?
    
    \blanklines{5}

  \item immigration/emigration at a constant per-capita rate?

    \blanklines{5}

  \item negative per-capita growth rate when the population is small, potentially due to difficulty finding partners? \hlwarn{Edited: ``to find'' \(\to\) ``finding''; ``the model account'' \(\to\) ``accounts'' (below)}  Let's say that if the population falls below a constant \(A\), then the population should shrink. We don't care about how it shrinks as long as the model accounts for the shrinkage. Moreover, if the population is exactly \(A\), then it should stay constant.

    \blanklines{15}
\end{enumerate}

\end{document}
