%! TeX program = lualatex
\documentclass[../main.tex]{subfiles}
\begin{document} \section{Simple models and qualitative information}

Before we investigate mathematical properties of differential equations, let's see why we should care about their properties at all by doing some real-world modelling. 

\begin{method} \label{method:diff-eq-model}
  Use differential equations to describe the rates of change of physical \hlmain{systems}.  The simplest idea is
  \begin{center}
    (rate of change) = (rate of internal change) + (rate of external change).
  \end{center}
\end{method}

\blanklines{10}

The phrase ``\emph{rate of change is \hlattn{proportional} to}'' is a good starting point to discuss rate of \hlmain{internal} change for many biological systems.  That \hlmain{the rate of change (with respect to time \(t\)) of some quantity \(P\)} is \hlattn{proportional} to \hlsupp{\(P\)} translates to a differential equation 
\[
  {\color{main}P'(t)} = {\color{attn} k} {\color{supp} P(t)},\quad \text{ for some constant } k.
\]

% Assume \(P \ge 0\).  The number \(k\) is called the \hlmain{proportionality constant}, and its sign reveals qualitative information.  Use our calculus knowledge to complete the table.
%
% \begin{table}[H]
%   \centering
%   \begin{tabular}{r|p{1.5in}|p{1.5in}|p{1.5in}}
%   & \(k < 0\) & \(k = 0\) & \(k > 0\) \\ \midrule
%     how \(P\) changes as \(t\) increases & & & \\[2ex]
%   \end{tabular}
% \end{table}

For the next three examples, we consider the interaction between a made-up sedation drug ABC and the human bloodstream in palliative scenario. % Eventually (next week), we would like to answer the question ``how much ABC will keep a patient sedated for \emph{blah} minutes?''
Assume that the human body does not naturally produce ABC, and ABC is not present in the human body unless it is administered externally. 
Assume the amount of ABC is measured in \emph{mg} and time in \emph{minutes}.

\begin{example}[modelling with only rate of internal change] \label{ex:diff-eq-ABCv1}
  Experimental data suggests that the \hlmain{rate of change of ABC} is \hlattn{proportional} to \hlmain{the amount of ABC present} with a proportionality constant of \(-1\%\) per minute.

  Write down a \emph{continuous-time model} of ABC present in the bloodstream.

  \blanklines{15}
\end{example}
\clearpage

\begin{example}[modelling with rates of internal and external change] \label{ex:diff-eq-ABCv2}
  Continue from Example~\ref{ex:diff-eq-ABCv1}.  Suppose further \(ABC\) is \emph{continuously} administered at a rate of \(r\) mg per minute. 

  Write down a \emph{continuous-time model} of ABC in the bloodstream.

  \blanklines{20}
\end{example}

\begin{exercise} \label{ex:diff-eq-ABCv3}
  Continue from Example~\ref{ex:diff-eq-ABCv2}.  Suppose further that \(ABC\) exits the bloodstream through blood loss at a rate of \(s\) mg (of ABC) per minutes. 

  Write down a \emph{continuous-time model} of ABC in the bloodstream.

  \blanklines{20}
\end{exercise}
\clearpage

Given enough information, we can pick up \hlmain{qualitative information} from these differential equations without solving it.  Recall that
\begin{enumerate}[itemsep={2ex}]
  \item if \(f'(t) > 0\), then \(f\) is \underline{\hspace{3in}}.
  \item if \(f'(t) = 0\), then \(f\) is \underline{\hspace{3in}}.
  \item if \(f'(t) < 0\), then \(f\) is \underline{\hspace{3in}}.
\end{enumerate}

\begin{example} \label{eg:DE-qualitative}
  Suppose a population model is \(\frac{dy}{dt} = ky\) for some unknown constant \(k\).  Describe how the population changes 
  \begin{enumerate}
    \item if \(k > 0\).
    \item if \(k = 0\).
    \item if \(k < 0\).
  \end{enumerate}
  
  \blanklines{10}
\end{example}

\begin{exercise}
  Suppose a differential equation is \(u'(t) u(t) = t u(t) - 1\). Describe whether its solution is increasing or decreasing at the specified time under the specified condition.
  \begin{enumerate}
    \item \(u(5) = 7\).
    \item \(u(-2) = 32\).
    \item \(u(17) = -1.47\).
  \end{enumerate}

  \blanklines{20}
\end{exercise}
\clearpage

Ultimately, a differential equation is a mathematical relation among several quantities.  As soon as information is available for all but one of them, we can determine information for the remaining unknown quantity.
\begin{example}
  Suppose a population is modelled by \(\frac{dN}{dt} = 1.13 N(1-N/5)\). If the population is increasing at \(t = 0\), what are the possible values for \(N(0)\)?

  \blanklines{20}
\end{example}

\begin{exercise}
  Suppose the temperature \(T(t)\) (measured in Celsius) of an object is modelled by
  \[
    \frac{dT}{dt} = \frac{12}{T(t) - 20} + 5 \frac{\sin(\pi t)}{t}.
  \]

  If the temperature is decreasing at time \(t = 1\), what are the possible values of \(T(1)\)?

  \blanklines{20}
\end{exercise}
\clearpage
\end{document}
