%! TeX program = lualatex
\documentclass[../main.tex]{subfiles}
\begin{document} \section{The Bernoulli substitution}

A substitution, also known as a change of variable, can often \hlmain{transform} an unfamiliar problem into a familiar one to which existing tools apply. 

\begin{method}[Bernoulli substitution] \label{method:bernoulli-sub}
  The \hlmain{Bernoulli substitution} \(u = y^{1-n}\) applies to \hlsupp{non-linear}, first-order differential equations \hlwarn{Edited: ``equation'' \(\to\) ``equations''} in the form
  \begin{equation} \label{eq:bernoulli}
    y'(t) + P(t) y(t) = Q(t) {\color{warn} \big(y(t)\big)^{n}}.
  \end{equation}

  This type of DE is also known as a Bernoulli differential equation. \hlwarn{Edited: ``Such a type of DEs is also known as Bernoulli differential equations'' \(\to\) ``This type of DE is also known as a Bernoulli differential equation''}
\end{method}
\faStar{} The substitution \(u = y^{1-n}\) requires us to substitute \(y'(t)\) in terms of \(u'(t)\). We can use \emph{implicit differentiation} to figure out how this works.

\blanks{10}

\begin{example}
  When should we use the Bernoulli substitution?

  Study the \emph{forms} of the following differential equations and determine the applicability of the Bernoulli substitution in each case. 
  
  \begin{enumerate}[wide, label=(\alph*)]
    \item \(v'(t) + (3tv(t))^{2} = 1\).

      \blanks{3}
    \item \(v'(t) + \frac{1}{v(t)} + t v(t) = 0\).

      \blanks{3}
    \item \(\cos(t) v'(t) + \sqrt{3t v(t)} = \sin(t)(v(t))^{3}\).
      \blanks{3}

  \end{enumerate}
\end{example}

\clearpage

Bernoulli substitution \emph{transforms} a non-linear, first-order differential equation to a linear one. It works, \emph{in spirit}, like Method~\ref{method:recurrence-sub} in the sense that we solve a simpler problem first and recover the solution later.
\begin{example} 
  Solve the initial-value problem \(y'(t) - \frac{2 y(t)}{t}  = -\big( ty(t) \big)^{2}\) with \(y(1) = 1/2\).

  \blanks{3}

  \begin{enumerate}[wide, label=Step~(\arabic*).]
    \item Identify the substitution \(u = \underline{\hspace{1in}}\) and calculate \(u'\).

      \blanks{5}

    \item Multiply the given Bernoulli DE through by \underline{\hspace{1in}} and perform the substitution to get a new DE.

      \blanks{10}

    \item Use the substitution to find the new initial condition \underline{\hspace{2in}}.

    \item Solve the new initial-value problem \underline{\hspace{3in}}.

      \blanks{5}

    \item Recover the solution of the original IVP by turning \(u(t)\) back into \(y(t)\).

      \blanks{10}
  \end{enumerate}
\end{example}
\clearpage

\begin{example}
  Solve the initial-value problem \(y'(t) + \frac{1}{1 - t} y(t)  = \sqrt[5]{y(t)}\) with \(y(2) = 0\).

  \blanks{50}
\end{example}
\clearpage

\begin{exercise}[concept check]
  Find the general solution to the logistic equation
  \[
    N'(t) = rN(t) \left( 1 - \frac{N(t)}{K} \right), \text{ where \(r,K\) are positive constants}.
  \]

  {\footnotesize Hint: Start by reorganizing the equation to the form in Equation~\eqref{eq:bernoulli}.}

  \blanks{50}
\end{exercise}
\clearpage

\begin{exercise}[Midterm, Winter 2025]
  Consider the differential equation \(\frac{dy}{dt} + \frac{6y}{t} = y^{2}\).

  \begin{enumerate}[wide]
    \item Perform a suitable substitution to reduce this to a linear differential equation. 

      \blanks{15}

    \item Solve the linear differential equation you got in part (a).

      \blanks{23}

    \item Find the general solution to the original differential equation.

      \blanks{10}
  \end{enumerate}
\end{exercise}
\clearpage
\end{document}
