%! TeX program = lualatex
\documentclass[../main.tex]{subfiles}
\begin{document} \section{Linearity}

We learn a bit more theory of vectors and matricies to learn some calculation useful for next week's material (and in general, but that's besides the point). 

An equation of the form \(A \vec{x} = \vec{b}\) takes vectors as inputs (left-hand side) and spits out vectors (right-hand side). This interpretation says matrix-vector multiplication is a function \(F(\vec{x}) = A \vec{x}\). Such functions are called vector-valued functions because their outputs are vectors.

\blanklines{10}

We learn to work with a particularly simple class of vector-valued functions called linear functions.

\begin{definition}[linearity] \label{def:linearity}
  A function \(f(\vec{x})\) is called \hlmain{linear} if \(F(\vec{x}) = A\vec{x}\) for some matrix \(A\). If a function is not a matrix-vector multiplication, then it is not linear. All linear functions satisfy the property
  \begin{equation} \label{eq:linearity}
    A(a \vec{u} + b \vec{v}) = a A\vec{u} + b A\vec{v} \quad\text{where \(a,b\) can be any constants.}
  \end{equation}
\end{definition}
\blanklines{5}

Equation~\eqref{eq:linearity} seems counter-intuitive because why perform two matrix-vector multiplications (right-hand side) when you can just do it once (left-hand side).  Matricies have special vectors, called \emph{eigenvectors}, for which matrix-vector multiplications are \emph{extremely} simple.  In \thecoursesubject{}~\thecoursenumb{}, we want to rewrite a vector in terms of eigenvectors and apply Equation~\ref{eq:linearity} to simplify our computation for \(A^{n} \vec{x}\). We will learn eigenvectors next week. For now, we learn to rewrite vectors in terms of eigenvectors of matricies.

\begin{method}[express a vector in terms of other vectors] \label{method:linear-combination}
  To find unknown constants \(a,b\) so that \(\vec{x} = a \vec{u} + b \vec{v}\) where \(\vec{x}, \vec{u}, \vec{v}\) are given, rewrite the equality as (usual) linear equations involving unknown \(a,b\), then use back substitution.
\end{method}

\faStar{} In \thecoursesubject{}~\thecoursenumb{}, Method~\ref{method:linear-combination} is often used in conjunction with Equation~\ref{eq:linearity}. However, they are not married to each other in general. 

{\footnotesize \faExclamationTriangle{} In \thecoursesubject{}~\thecoursenumb{}, such problems will always have solutions, because we only deal with ``special enough'' vectors \(\vec{u}, \vec{v}\). However, such equations need not have a solution if \(\vec{u}, \vec{v}\) are not ``special enough.'' For those interested, the linear algebraic concept for ``special enough'' is called \emph{linear independence}.}
\clearpage

Consider a matrix \(A = \begin{bmatrix} 0 & 2 \\ 1 & 1 \end{bmatrix}\) and its two eigenvectors \(\vec{u} = \begin{bmatrix} 1 \\ 2 \end{bmatrix}\) and \(\vec{v} = \begin{bmatrix} 1 \\ 0 \end{bmatrix}\).  Our goal is to find \hlmain{a simple answer} for arbitrary powers \(A^{n} \vec{x}\) where \(\vec{x} = \begin{bmatrix} -11 \\ 10 \end{bmatrix}\).

\begin{example} \label{ex:linear-combination}
  Find unknown constants \(a,b\) so that \(\vec{x} = a \vec{u} + b \vec{v}\). Hint: Use Method~\ref{method:linear-combination}.

  \blanklines{20}
\end{example}

\begin{example}
  Calculate \(A^{n} \vec{u}\) and \(A^{n} \vec{v}\) for arbitrary positive integers \(n\). 

  \blanklines{15}
\end{example}

\begin{example}
  Calculate \(A^{n} \vec{x}\) for arbitrary positive integers \(n\). Hint: Use Equation~\eqref{eq:linearity}.

  \blanklines{5}
\end{example}
\clearpage

\end{document}
