%! TeX program = lualatex
\documentclass[../main.tex]{subfiles}
\begin{document} \section{Expectation and variance}

Probability gives a big picture view of data. Pictorially, probability mass/density functions let us visualize the data. Other common ways of thinking about data involves computing their \hlmain{expectation} and \hlmain{variance}. 

\begin{definition}[expectation]
  The \hlmain{expectation} of a discrete random variable \(X\) with PMF \(f_{X}(x)\) and values \(x_{1}, \ldots, x_{n}\) is
  \begin{equation} \label{eq:expectation-discrete}
    E(X) = \sum_{i = 1}^{n} x_{i} f_{X}(x_{i}).
  \end{equation}

  The \hlmain{expectation} of a continuous random variable \(X\) with PDF \(f(x)\) is
  \begin{equation} \label{eq:expectation-continuous}
    E(X) = \int_{-\infty}^{\infty} x f(x) \;dx.
  \end{equation}
\end{definition}

\begin{example} \label{ex:variance-discrete-betting}
  We play a betting game of predicting the outcome of rolling a fair six-sided die.  Each bet costs \(1\) chore token. If we win, we collect \(2\) chore tokens.

  This game is modelled by the random variable \(X\) where \(X = \underline{\hspace{1cm}}\) is the event in which we lose and \(X = \underline{\hspace{1cm}}\) is the event in which we win.  We work out that \(\mathbb{P}(X = \underline{\hspace{1cm}}) = \underline{\hspace{1cm}}\) and \(\mathbb{P}(X = \underline{\hspace{1cm}}) = \underline{\hspace{1cm}}\).  What is the expected return if we play the game \(10\) times?

  \blanklines{10}
\end{example}

\begin{example} \label{ex:variance-continuous-uniform}
  Let \(X\) be a continuous random variable with PDF \(f(x) = 1/5\) on \([-1, 1]\), \(f(x) = 1/10\) on \([2,8]\) and \(0\) everywhere else.  Calculate \(E(X)\).

  \blanklines{15}
\end{example}
\clearpage

Variance is defined as \(Var(X) = E( (X - E(x))^{2} )\) regardless of whether \(X\) is discrete or continuous. However, this formula can be specialized.  
\begin{definition}[variance]
  Let \(\mu = E(x)\). Recall that \(\mu\) is a constant.

  The \hlmain{variance} of a discrete random variable \(X\) with PMF \(f_{X}(x)\) and values \(x_{1}, \ldots, x_{n}\) is 
  \begin{equation} \label{def:variance-discrete}
    Var(X) = \sum_{i=1}^{n} (x_{i} - \mu)^{2} f_{X}(x_{i}).
  \end{equation}

  The \hlmain{variance} of a continuous random variable \(X\) with PDF \(f(x)\) is 
  \begin{equation} \label{def:variance-continuous}
    Var(X) = \int_{-\infty}^{\infty} (x - \mu)^{2} f(x) \;dx.
  \end{equation}
\end{definition}

\begin{example}
  Calculate the variance for the discrete random variable in Example~\ref{ex:variance-discrete-betting}.

  \blanklines{10}
\end{example}

\begin{example}
  Calculate the variance for the continuous random variable in Example~\ref{ex:variance-continuous-uniform}.

  \blanklines{15}
\end{example}
\clearpage

What do expectation and variance measure?
\end{document}
