%! TeX program = lualatex
\documentclass[../main.tex]{subfiles}
\begin{document} \section{The law of total probability and independence}

The law of total probability mathematically takes advantage \hlwarn{Edited: ``takes advantages of'' \(\to\) ``takes advantage of''} of mutually exclusive and exhaustive events to \hlmain{relate} probabilities.

First, let's express \(\mathbb{P}(E \cap F)\) in terms of \(\mathbb{P}(E \mid F)\).

\blanks{10}

If we divide the sample space into mutually exclusive and exhaustive events \(F_{1}\) and \(F_{2}\) and try to compute \(\mathbb{P}(E)\), then we should get ...

\blanks{20}

In the abstract, we have ...

\blanks*

\clearpage

If we divide the sample space into mutually exclusive and exhaustive events \(F_{1}, F_{2}, F_{3}\) and try to compute \(\mathbb{P}(E)\), then we should get ...

\blanks{10}

\faStar{} The law of total probability is the above pattern in full generality. Suppose \(F_{1}, \ldots, F_{n}\) are mutually exclusive and exhaustive events in a sample space \(\Omega\). The \hlmain{law of total probability} is a way to express \(\mathbb{P}(E)\) as ...

\blanks{10}

\begin{example}
  Suppose we have a bag of red, green and blue cubes and spheres with the following probability data.  
  \[
    \mathbb{P}(\text{red}) = 30\%, \quad
    \mathbb{P}(\text{green}) = 60\%, \quad
    \mathbb{P}(\text{blue}) = 10\%,\quad
    \mathbb{P}(\text{red cubes}) = 5\%,\quad
    \mathbb{P}(\text{cube} \mid \text{blue}) = 8\%.
  \]
  What is the probability that a randomly chosen object is a green cube?

  \blanks*
\end{example}
\clearpage

Now we turn to independence. Intuitively, two events are independent if the occurrence of one \hlmain{does not affect} the other.  

\blanks{15}

\begin{definition}[independent events]
  Two events \(E,F\) in a common sample space are said to be \hlmain{independent} if 
  \begin{equation} \label{eq:probability-independence}
    \mathbb{P}(E \mid F) = \mathbb{P}(E)
    \quad\text{or, equivalently,}\quad
    \mathbb{P}(E \cap F) = \mathbb{P}(E) \mathbb{P}(F).
  \end{equation}
\end{definition}

To help you understand these equivalent definitions, we deduce them from each other.

\blanks{15}

\begin{example}
  Let \(E, F\) be events in a common sample space. Let \(\mathbb{P}(E) = 0.3\), \(\mathbb{P}(F) = 0.2\) and \(\mathbb{P}(E \cap F) = 0.06\). \hlwarn{Edited: ``\(\mathbb{P}(B)\)'' \(\to\) ``\(\mathbb{P}(F)\)'' and ``\(\mathbb{P}(A \cap B)\)'' \(\to\) ``\(\mathbb{P}(E \cap F)\)'' to match the events \(E,F\)} \hlwarn{FIXME: verify this math/content correction} Determine if \(E,F\) are independent.

  \blanks*
\end{example}
\clearpage

\begin{example}
  Let \(E = \{1,3\}, F = \{2,4\}\) be events in a common sample space \(\Omega = \{1,2,3,4,5,6\}\).  Are \(E,F\) mutually exclusive? Are \(E,F\) independent?

  \blanks{10}
\end{example}

\begin{exercise}
  If the event \(E \cap F\) is empty, are the events \(E,F\) independent?

  \blanks{10}
\end{exercise}

\begin{exercise} 
  Let \(E,F\) be two events in a sample space \(\Omega\). If \(E,F\) are independent events and \(\mathbb{P}(E) > 0\), does that imply \(E\) and \(F\) are mutually exclusive?

  \blanks{10} 
\end{exercise}

\begin{exercise}
  If \(E\) and \(F\) are independent events, does that mean \(E^{c}\) and \(F\) are also independent events?

  \blanks{10}
\end{exercise}
\end{document}
