%! TeX program = lualatex
\documentclass[../main.tex]{subfiles}
\begin{document} \section{Mathematical techniques for recursive models}

Now that we know how to turn words into recursive models, we learn techniques to investigate recurrence.

At the end of this section, we will be able to do the following.
\begin{enumerate}
  \item Verify proposed solutions of recursions.
  \item Understand the effect of initial conditions and find equilibrium solutions.
  \item Solve simple recursions.
  \item Use calculus to solve problems involving discrete-time models.
\end{enumerate}

\faLightbulb{} Although Example 2.5 is not a biological model, it is intuitive and sophisticated enough to demonstrate all of this section's learning goals.

To calculate a few terms of a given recurrence, we simply start with time \(t=1\) and calculate \(u_{1}\) using the recurrence equation. Once \(u_{1}\) is calculated, we apply \(t = 2\) to the recurrence equation to calculate \(u_{2}\) and so on.
\begin{example} \label{ex:recurrence-bruteforce}
  Calculate the balance at the end of years 1, 2, and 3 using the model in Example~\ref{ex:recurrsion-investment-model}.

  \blanklines{20}
\end{example}

\begin{exercise} \label{ex:recurrence-equilibrium}
  Calculate the balance at the end of years \(3\) using the model in Example~\ref{ex:recurrsion-investment-model} but change the initial deposit to \(\$ 2,500\).

  \blanklines{10}
\end{exercise}

\clearpage

Theoretically, we can always use the recursion equation to calculate \(u_{t}\) for any arbitrary \(t\), say \(t = 10^{30}\), of a recursion. However, this approach is time consuming (even for a computer if a recursion is sufficiently complicated). Moreover, this approach does not give us any insight into the behaviour of the model. We can't really tell if a model follows a power rule, is exponential or something else completely. 

In some cases, we can use mathematical techniques to ``get rid'' of the recursion equation and directly expression \(u\) as a function of \(t\). This is called \hlmain{solving a recursion}.

First, let's understand what it means for a function to be a solution of a recursion. 

\begin{definition}
  Suppose we have a recurrence \(u_{t}\) with some initial condition \(u_{0}\). A function \(f(t)\) is a solution of the recursion if 
  \begin{enumerate}
    \item \(f(t)\) satisfies the defining recursion equation for \(u_{t}\), \emph{and}
    \item \(u_{0} = f(0)\).
  \end{enumerate}

  Note that \(f(t)\) cannot refer to any \(u_{t}\) and must be a well-defined function of \(t\).
\end{definition}

\begin{example}
  Is \(a_{t} = 2^{2^{t-1}}\) the solution for \(a_{1} = 2\) and \(a_{t} = a_{t-1}^{2}\) for \(t \ge 2\)?

  \blanklines{30}
\end{example}
\clearpage

\begin{exercise}[concept check and algebra skill practice]
  Is \(b_{t} = -1500 (1.02)^{t} + 2500\) is a solution of the recurrence in Example~\ref{ex:recurrsion-investment-model}?

  \blanklines{50}
\end{exercise}
\clearpage

Guess-and-check is an inefficient method of looking for solutions of recursions. We are better off looking for systematic methods.

\begin{method} \label{method:recurrence-exp}
  Let \(a > 0\) be a constant. The solution to the recurrence \(u_{t} = a u_{t-1}\) is 
  \[
    u_{t} = u_{0} a^{t}.
  \]

  In this statement, \(u_{0}\) can be any constant.
\end{method}

Why does this work?

\blanklines{5}

\faStar{} Recurrence offers a convenient language to develop discrete-time models. However, none of the calculus tools apply \emph{directly} to recurrence. \hlmain{Solving a recurrence converts a unfamiliar problem to a familiar problem.}

\begin{example}
  Suppose a population grows at rate \(r\) (annually) with an initial population \(P_{0} = 1,000\).   What is the long-term behaviour of the population , i.e., \(\lim_{t \to \infty} P_{t}\), when
  \begin{enumerate}
    \item \(r = 1\)?
    \item \(r > 1\)?
    \item \(0 < r < 1\)?
  \end{enumerate}

  \blanklines{25}
\end{example}
\clearpage


\clearpage

Recurrences are sensitive to their initial condition. Consider once again Example~\ref{ex:recurrsion-investment-model}. Recall our savings account pays \(2\%\) annual interest rate, and we withdrawal \(50\) dollars after each interest payment.  Initial deposits of \(1,000\) vs \(2,500\) should lead to different balances after \(10\) years.  This is true for all recurrences: Different initial conditions lead to different solutions.

However, the initial deposit \(2,500\) is special as we saw in Exercise~\ref{ex:recurrence-equilibrium} because it leads to no change from year to year.

\begin{definition}[equilibrium solution]
  An equilibrium solution of a recurrence equation (not the whole recurrence) is a constant function.  We denote equilibrium solutions by making the variable wear a hat and throwing away subscripts, e.g., \(\hat{u}\).
\end{definition}

\blanklines{10}

\faStar{} Equilibrium solutions are very easy to find because they are constants. Simply replace all \(u_{t}\) by \(\hat{u}\) and solve for \(\hat{u}\) using basic algebra. It is possible to find more than one equilibrium solutions.

\begin{example}
  Find the equilibrium solution for the recurrence equation \(v_{t} = 1.5 v_{t-1} + 35\).

  \blanklines{25}
\end{example}
\clearpage

\begin{exercise}[concept check]
  Find \emph{all} equilibria of the recurrence equation \[u_{t} = \sqrt{3 u_{t-1} - 2}.\]

  \blanklines{20}
\end{exercise}

\begin{exercise}[concept check]
  Find equilibria (one or many?) of the recurrence equation \[p_{t} = \frac{1}{2} p_{t-1} - \frac{1}{2}.\]

  \blanklines{20}
\end{exercise}
\clearpage

Now we advance to slightly more complicated recurrence. 

\begin{method} \label{method:recurrence-sub}
  Consider a recurrence of the form \(v_{t} = A v_{t-1} + B\) whose equilibrium is always \(\hat{v} = \frac{B}{1 - A}\). Introduce a new recurrence \(u_{t}\) and perform the change of variable \(v_{t} = u_{t} + \hat{v}\). Solve the new recurrence \(u_{t}\) and recover the original solution.
\end{method}

\begin{example}
  Solve the recurrence \(b_{t} = 1.02 b_{t-1} - 50\) with initial condition \(b_{0} = 1000\).

  \begin{enumerate}[wide, label={Step~(\arabic*).}]
    \item Perform the change of variable.

      \blanklines{25}

    \item Calculate \(u_{0}\) and solve for the new recurrence in \(u_{t}\). 

      \blanklines{10}

    \item Recover the solution of the original recurrence.

      \blanklines{5}
  \end{enumerate}
\end{example}
\clearpage

The substitution process in Method~\ref{method:recurrence-sub} can be packaged into a formula.
\begin{method}
  The solution to the recursion \(v_{t} = A v_{t-1} + B\) with initial condition \(v_{0}\) is
  \[
    v_{t} = (v_{0} - \hat{v}) A^{t} + \hat{v}, \text{ where } \hat{v} = \frac{B}{1 - A}.
  \]
\end{method}

\begin{exercise}[concept check]
  Solve the recurrence \(p_{0} = 1\) and \(p_{t} = \frac{1}{2} p_{t-1} - \frac{1}{2}\) for integers \(t \ge 1\) using the formula and the substitution method. Make sure you understand both.

  \blanklines{40}
\end{exercise}

\clearpage

Solutions of recurrence allow us to answer interesting questions.  This is the last learning objective of this week.
\begin{example}
  Consider the recurrence in Example~\ref{ex:recurrsion-investment-model} 
  \[
    b_{0} = 1000 \quad\text{and}\quad b_{t} = 1.02 b_{t-1} - 50, \text{ for integers } t \ge 1. 
  \] 
  Does the balance ever reach \(0\)? If so, roughly how many years since the initial investment should we expect the balance to be \(0\)? 
  
  Don't simply list all \(b_{t}\). Use the solution of the recurrence and calculus to solve this problem.

  \blanklines{45}
\end{example}
\clearpage

\begin{exercise}
  Let's modify Example~\ref{ex:recurrsion-investment-model} a bit further and assume the initial investment is an unknown.  Everything else remains unchanged.  

  What is the initial investment if the money were to be depleted after \(10\) years?

  \blanklines{45}
\end{exercise}
\clearpage

Here is a brain teaser.

\begin{exercise}[challenge of the week]
  A house has a cat and some mouse. The cat chases away a fixed portion of the mouse everyday during the night.  Voluntary move-ins/outs happen before midnight. A census of the mouse population is taken at midnight.

  An intern proposes the following model for the number of mouse \(m_{t}\) at night \(t\). 
  \[
    m_{0} \text{ is unknown} \quad\text{and}\quad m_{t} = 0.2 m_{t-1} + 5, \text{ for \(t \ge 1\)}.
  \]

  According to the model, what \emph{percentage} of the mouse population is chased away by the cat everyday?

  {\footnotesize Hint: Exercise~\ref{ex:emc2}. The challenge is to figure out what we learned over the past two weeks is useful.}
\end{exercise}
\end{document}
