%! TeX program = lualatex
\documentclass[../main.tex]{subfiles}

\newenvironment{solution}{
  \begin{proof}[Sample solution]
    \color{main}
  }{
  \end{proof}
}

\begin{document}

\chapter*{Week 2 practice problems}

\begin{exercise}
  For each recurrence below, calculate \(u_{1}, u_{2}, u_{3}\).
  \begin{enumerate}
    \item \(u_{0} = 3\) and \(u_{t} = u_{t-1}^{2}\) for integers \(t \ge 1\)

      \begin{solution}
        \begin{align*}
          u_{0} &= 3, &\text{(given)} \\
          u_{1} &= u_{0}^{2} = 3^{2} = 9, \\
          u_{2} &= u_{1}^{2} = 9^{2} = 81,\\
          u_{3} &= u_{2}^{2} = 81^{2}. 
        \end{align*}
      \end{solution}

    \item \(u_{0} = 1\) and \(u_{t} = \frac{1}{3} u_{t-1} + \frac{1}{4}\) for integers \(t \ge 1\).

      \begin{solution}
        \begin{align*}
          u_{0} &= 1, &\text{(given)} \\
          u_{1} &= \frac{1}{3}u_{0} + \frac{1}{4} = \frac{1}{3} \cdot 1 + \frac{1}{4} = \frac{7}{12}, \\
          u_{2} &= \frac{1}{3}u_{1} + \frac{1}{4} = \frac{1}{3} \cdot \frac{7}{12} + \frac{1}{4} = \frac{4}{9}, \\
          u_{3} &= \frac{1}{3}u_{2} + \frac{1}{4} = \frac{1}{3} \cdot \frac{4}{9} + \frac{1}{4}.
        \end{align*}
      \end{solution}

    \item \(u_{0} = 8\) and \(u_{t} = \frac{1}{2} u_{t-1}\).
  \end{enumerate}
\end{exercise}


\begin{exercise}
  Determine the long-term behaviour of these recurrences.

  \begin{enumerate}
    \item \(u_{0} = 3\) and \(u_{t} = 1.1 u_{t-1}\) for integers \(t \ge 1\).

      \begin{solution}
        The solution to the given recurrence is \(u_{t} = 3 \cdot 1.1^{t}\).  Its long term behaviour is 
        \[
          \lim_{t \to \infty} u_{t} = \lim_{t \to \infty} 3 \cdot 1.1^{t} = \infty.
        \]
        It means that \(u_{t}\) will increase without bound as \(t\) increases.
      \end{solution}

    \item \(u_{0} > 0\) and \(u_{t} = \frac{1}{2} u_{t-1}\) for integers \(t \ge 1\).

      \begin{solution}
        The solution to the given recurrence is \(u_{t} = u_{0} \cdot \left(\frac{1}{2}\right)^{t}\).  Its long term behaviour is 
        \[
          \lim_{t \to \infty} u_{t} = \lim_{t \to \infty} u_{0} \cdot \left( \frac{1}{2} \right)^{t}  = u_{0} \lim_{t \to \infty} e^{\ln(1/2)t} = \lim_{t \to \infty} e^{(\text{negative constant}) t} = 0.
        \]
        It means that \(u_{t}\) will decrease to \(0\) as \(t\) increases.
      \end{solution}

    \item \(u_{0} < 0\) and \(u_{t} = \frac{3}{4} u_{t-1}\) for integers \(t \ge 1\).

      \begin{solution}
        The solution to the given recurrence is \(u_{t} = u_{0} \cdot \left(\frac{3}{4}\right)^{t}\).  Its long term behaviour is 
        \[
          \lim_{t \to \infty} u_{t} = \lim_{t \to \infty} u_{0} \cdot \left( \frac{3}{4} \right)^{t}  = u_{0} \lim_{t \to \infty} e^{\ln(3/4)t} = \lim_{t \to \infty} e^{(\text{negative constant}) t} = 0.
        \]
        It means that \(u_{t}\) will increase to \(0\) as \(t\) increases.
      \end{solution}

    \item \(u_{0} = -1\) and \(u_{t} = \sqrt{2} \cdot u_{t-1}\) for integers \(t \ge 1\).

      \begin{solution}
        The solution to the given recurrence is \(u_{t} = - \left(\sqrt{2} \right)^{t}\).  Its long term behaviour is 
        \[
          \lim_{t \to \infty} u_{t} = \lim_{t \to \infty} -\left( \sqrt{2} \right)^{t} = - \lim_{t \to \infty} e^{\ln(\sqrt{2})t} = -\lim_{t \to \infty} e^{(\text{positive constant}) t} = -\infty.
        \]
        It means that \(u_{t}\) will decrease to \(-\infty\) as \(t\) increases.
      \end{solution}
  \end{enumerate}
\end{exercise}

% \begin{exercise}
%   Which of the following describe an equilibrium solution of an unknown recurrence equation involving \(u_{t}\)?
%
%   \begin{enumerate}
%     \item \(u_{t} = 0\) for every \(t \ge 0\).
%     \item \(u_{t}\) is constant with respect to \(t\).
%     \item \(u_{0} = u_{1} = u_{2} = u_{3} = \cdots\) and so on.
%     \item \(u_{t} = t\) for every \(t \ge 0\).
%   \end{enumerate}
% \end{exercise}

\begin{exercise}
  For each recurrence equation below, find their equilibria.
  \begin{enumerate}
    \item \(2u_{t} = u_{t - 1}^{2} + 1\).
      \begin{solution}
        Let \(\hat{u}\) denote the equilibrium solution which is just a constant that satisfies the recurrence.  Replace every \(u_{t}\) and \(u_{t-1}\) by \(\hat{u}\) and solve for it.  We get
        \begin{align*}
          2\hat{u} &= \hat{u}^{2} + 1 \\
          \hat{u}^{2} - 2\hat{u} + 1 &= 0 \\
          \hat{u} = 1
        \end{align*}
        There is only one equilibrium \(\hat{u} = 1\).

      \end{solution}
    \item \(w_{t} = 3\sqrt{w_{t - 1}w_{t - 2} - 1}\).
      \begin{solution}
        Let \(\hat{w}\) denote the equilibrium solution which is just a constant that satisfies the recurrence.  Replace every \(w_{t}, w_{t-1}, w_{t-2}\) by \(\hat{w}\) and solve for it.  We get
        \begin{align*}
          \hat{w} &= 3\sqrt{\hat{w}^{2} - 1} \\
          \hat{w}^{2} &= 9(\hat{w}^{2} - 1) \\
          \hat{w} &= \pm \sqrt{\frac{9}{8}}.
        \end{align*}
        There are two equilibrium solutions \(\hat{w} = \sqrt{\frac{9}{8}}\) and \(\hat{w} = -\sqrt{\frac{9}{8}}\).
      \end{solution}

    \item \(p_{t} = \frac{2}{3}p_{t-1}\).

      \begin{solution}
        Let \(\hat{p}\) denote the equilibrium solution which is just a constant that satisfies the recurrence.  Replace every \(p_{t}, p_{t-1}\) by \(\hat{p}\) and solve for it.  We get
        \begin{align*}
          \hat{p} &= \frac{2}{3}\hat{p} \\
          \hat{p} &= 0.
        \end{align*}
        The only equilibrium is \(\hat{p} = 0\).
      \end{solution}

    \item \(p_{t} = \frac{2}{3}p_{t-1} - \frac{1}{5}\).

      \begin{solution}
        Let \(\hat{p}\) denote the equilibrium solution which is just a constant that satisfies the recurrence.  Replace every \(p_{t}, p_{t-1}\) by \(\hat{p}\) and solve for it.  We get
        \begin{align*}
          \hat{p} &= \frac{2}{3}\hat{p} - \frac{1}{5} \\
          \hat{p} &= \frac{5}{3}.
        \end{align*}
        The only equilibrium is \(\hat{p} = \frac{5}{3}\).
      \end{solution}

    \item \(w_{t} = w_{t-1} - \frac{1}{5}\).
      \begin{solution}
        Let \(\hat{w}\) denote the equilibrium solution which is just a constant that satisfies the recurrence.  Replace every \(w_{t}, w_{t-1}\) by \(\hat{w}\) and solve for it.  We get
        \begin{align*}
          \hat{w} &= \hat{w} - \frac{1}{5} \\
          0 &= -\frac{1}{5}.
        \end{align*}
        This is not possible. Therefore, the equilibrium solution does not exist.
      \end{solution}
  \end{enumerate}
\end{exercise}

\begin{exercise}
  How is an equilibrium different from the long-term behaviour of a recurrence?

  \begin{solution}
    Let's say the symbol (without the subscript) of our recurrence is \(w\).

    \begin{itemize}
      \item An equilibrium solution is just a constant that satisfies the recurrence. In symbolic form, an equilibrium means \(w_{0} = w_{1} = w_{2} = \cdots\) forever.
      \item The long-term behaviour of a recurrence is the limit \(\lim_{t \to \infty} w_{t}\). To computer this limit, we need\footnote{There are other ways to do it without solving it. But that's outside the scope of this course.} to solve the recursion.
    \end{itemize}
  \end{solution}
\end{exercise}

\begin{exercise}
  Solve the following recurrence.
  \begin{enumerate}
    \item \(v_{0} = 1\) and \(v_{t} = \frac{1}{3} v_{t-1} + \frac{1}{4}\).
      \begin{solution}
        Notice that \(a = 1/3\) and \(b = 1/4\). Because \(a \ne 1\), we have an equilibrium \(\hat{v} = \frac{b}{1-a} = \frac{1/4}{2/3} = \frac{3}{2}\).  

        Depending on your preference, either pull out the formula \(v_{t} = (v_{0} - \hat{v}) a^{t} + \hat{v}\) or go through the substitution \(u_{t} = v_{t} + \hat{v}\).

        \begin{itemize}
          \item If you use the formula, then \(v_{t} = \left(1 - \frac{3}{2}\right) \left( \frac{1}{3} \right)^{t} + \frac{3}{2}\). You are done.

          \item If you substitute \(u_{t} = v_{t} + \hat{p}\), then the recurrence becomes
            \[
              u_{0} = \left(1 - \frac{2}{3}\right) \quad\text{and}\quad u_{t} = \left(1 - \frac{3}{2}\right) u_{t-1}.
            \]
            Solve this recurrence to get \(u_{t} =\left(1 - \frac{2}{3}\right) \left(\frac{1}{3}\right)^{t}\). Put back \(v_{t}\) to get \(v_{t} = \left(1 - \frac{3}{2}\right) \left( \frac{1}{3} \right)^{t} + \frac{3}{2}\). It's the same solution as above.
        \end{itemize}
      \end{solution}

    \item \(u_{0} = 8\) and \(u_{t} = \frac{1}{2} u_{t-1}\).
      \begin{solution}
        There isn't much to do except straight up apply the formula \(u_{t} = u_{0} a^{t}\) where \(a = \frac{1}{2}\). Therefore, 
        \[
          u_{t} = 8 \left(\frac{1}{2}\right)^{t}.
        \]

      \end{solution}
    \item \(v_{0} = 8\) and \(v_{t} = \frac{1}{2} v_{t-1} + \frac{1}{2}\).
      \begin{solution}
        Notice that \(a = 1/2\) and \(b = 1/2\). Because \(a \ne 1\), we have an equilibrium \(\hat{v} = \frac{b}{1-a} = \frac{1/2}{1/2} = 1\).  Applying the formula quickly gives us the answer \(v_{t} = 7 \left(\frac{1}{2}\right)^{t} + 1\).
      \end{solution}
      
    \item \(b_{0} = 300\) and \(b_{t} = 1.2 b_{t-1} - 10\).
      \begin{solution}
        Notice that \(A = 1.2\) and \(B = -10\). Don't forget the mimes sign. Because \(A \ne 1\), we have an equilibrium \(\hat{b} = \frac{B}{1-A} = \frac{-10}{-1.2} = 50\).  Applying the formula quickly gives us the answer \(b_{t} = 250 \left(1.2\right)^{t} + 50\).

        The solutions make sense if you think of the \(1.2\) as annual return on investment and the \(-10\) as annual withdrawal. Our initial investment \(300\) earns \(300 \times 20\% = 60\) interest at the end of the first year, but we only take out \(10\). So the principle grows at the end of each year.
      \end{solution}

    \item \(a_{0} = 1000\) and \(a_{t} = 1.15 a_{t-1} + 10\).
      \begin{solution}
        Notice that \(A = 1.15\) and \(B = 10\). Because \(A \ne 1\), we have an equilibrium \(\hat{a} = \frac{B}{1-A} = \frac{10}{1 - 1.15} = -\frac{10}{0.15}\).  Applying the formula quickly gives us the answer \(v_{t} = \left(1000 - \frac{10}{0.15}\right) \left(1.15\right)^{t} - \left( \frac{10}{0.15} \right)\).
      \end{solution}

  \end{enumerate}
\end{exercise}

\begin{exercise}
  For each recurrence below, find the unknown parameter so that the given condition is satisfied.
  \begin{enumerate}
    \item \(m_{t} = m_{t-1} + 30\) for integers \(t \ge 1\).  Find the unknown \(m_{0}\) so that \(m_{30} = 2750\).
    \item \(b_{0} = 3000\) and \(b_{t} = a m_{t-1} - 100\) for integers \(t \ge 1\).  Find the unknown constant \(a\) so that \(b_{17} = 1000\).
    \item \(p_{0} = 1\) and \(p_{t} = \frac{1}{3} m_{t-1} + c\) for integers \(t \ge 1\).  Find the unknown \(r\) so that \(b_{17} = 1000\).
  \end{enumerate}
\end{exercise}

\begin{exercise}
  A radioactive substance has a half-life of \(300\) years.   If the amount of substance is \(10\) grams, how many years will it take for the substance to decay to \(0.1\) milligrams?
\end{exercise}

\begin{exercise}
  Carbon-14 has a half-life of \(5,730\) years. If an artifact has \(7\) grams of Carbon-14 initially and has \(5\) gram left, estimate the age of the artifact.
\end{exercise}

\begin{exercise}
  Suppose a person retires on January 1, 2065.

  Suppose they have one million dollars at the beginning of retirement.  The fund is invested and returns \(5\%\) interest on December 31 every year. Suppose they withdraw \(50,000\) at the beginning of each year for annual expenses, starting at retirement.

  Will this person ever run out of money? If so, when?
\end{exercise}

\begin{exercise}
  Someone plans to retire on January 1 some time in the future.

  Suppose their expenses is \(80,000\) per year withdrawn at the beginning of each year from their retirement fund. Moreover, suppose the retirement fund is invested and returns \(7\%\) interest on December 31 every year. 

  What is the minimum amount of money needed at the beginning of the retirement if the person is expected to live \(25\) years past retirement?
\end{exercise}
\end{document}
