
%! TeX program = lualatex
\documentclass[../main.tex]{subfiles}

\begin{document}

\section{Week 10 practice problems}

\begin{exercise}
  The sample mean and sample standard deviation of last year's final exam were \(70\) and \(12\), respectively.  This year, the mean and standard deviation for the final exam is \(72\) and \(14\). 

  \begin{enumerate}
    \item Which year's students performed better on the final exam?
    \item If A scored \(65\) on this year's final exam, what would be a comparable grade from last year's final exam?
  \end{enumerate}

  \begin{solution}
    This question is about the \(z\)-score of sample data.

    \begin{enumerate}
      \item They are very close. There is no definitive evidence that one performance is better than the other. Most of the test scores are near \(70\) to \(72\). This year's scores are slightly closer to \(70\), but last year's scores are slightly further away from \(72\). 

      \item A score of \(65\) this year has a \(z\)-score of \((65 - 72)/14 = -7/14 = -1/2\).  Therefore, the question asks ``what midterm score from last year has a \(z\)-score of \(-1/2\)?'' That is to solve for \(x\) in \((x - 70)/12 = -1/2\), which is \(x = 64\).
    \end{enumerate}
  \end{solution}
\end{exercise}

\begin{exercise}
  Suppose we have a set of sample data with standard deviation \(1.37\). If the \(z\)-score of a measurement (value - sample mean) is \(-2.5\).  What is the actual value of the measurement if the sample mean is \(3\)?

  \begin{solution}
    The value, say \(x\), must satisfy \((x - 3) / 1.37 = -2.5\). Therefore, \(x = -0.425\).
  \end{solution}
\end{exercise}

\begin{exercise}
  Let \(\vec{x}(t)\) be the number of juveniles and adults, respectively, in a population at time \(t\) in years. Every adult in the population produces, on average, \(2\) juveniles in the population next year. Every adult in the population survives to the next year with probability \(1/3\). Every juvenile survives to be an adult next year with probability \(1/2\).

  Develop a discrete-time model for the population.  Find \(\vec{x}(3)\).

  \begin{solution}
    Let \(x_{1}(t)\) represents the number of juveniles at time \(t\), and \(x_{2}(t)\) the number of adults at time \(t\).  Time is measured is years.  We have
    \[
      \begin{array}{rcrcrcr}
        x_{1}(t+1) &=&  & & 2 x_{2}(t) \\[2ex]
        x_{2}(t+1) &=& \frac{1}{2}x_{1}(t) &+& \frac{1}{3} x_{2}(t)
      \end{array}
    \]

    We can also express this as \(\vec{x}(t+1) = A \vec{x}(t)\) where \(A = \begin{bmatrix} 0 & 2 \\ 1/2 & 1/3 \end{bmatrix}\).  The initial condition is not given. Therefore, we can express \(\vec{x}(3)\) abstractly as \(\vec{x}(3) = A^{3} \vec{x}(0)\).
  \end{solution}
\end{exercise}

\begin{exercise}
  Perform the following vector operations pictorially.  Verify that your sketch is correct by performing the operations algebraically.

  \begin{itemize}
    \item \(-3 \begin{bmatrix} 3 \\ 1 \end{bmatrix}\).
    \item \(-2 \begin{bmatrix} 2 \\ -3 \end{bmatrix}\).
    \item \(\begin{bmatrix} -3/2 \\ -4 \end{bmatrix} + \begin{bmatrix} 3 \\ -1/2 \end{bmatrix}\).
    \item \(\begin{bmatrix} 1 \\ -2 \end{bmatrix} - \begin{bmatrix} 2 \\ 3 \end{bmatrix}\).
  \end{itemize}

  \begin{solution} 
    \phantom{top}

    \begin{tikzpicture}
      \begin{axis}[basic curve, black, xmin=-12, xmax=5, ymin=-5, ymax=3, xtick distance={2}, ytick distance={2}]
        \addplot coordinates {(0,0)};
        \coordinate (O) at (axis cs:0,0);
        \draw[-Latex, very thick] (O)--(axis cs:3,1) node[right] {\footnotesize \(\begin{bmatrix}3 \\ 1\end{bmatrix}\)};
        \draw[red, -Latex, very thick] (O)--(axis cs:-9,-3) node[left] {\footnotesize \(-3\begin{bmatrix}3 \\ 1\end{bmatrix}\)};
      \end{axis}
    \end{tikzpicture}
    \begin{tikzpicture}
      \begin{axis}[basic curve, black, xmin=-8, xmax=5, ymin=-5, ymax=7, xtick distance={2}, ytick distance={2}]
        \addplot coordinates {(0,0)};
        \coordinate (O) at (axis cs:0,0);
        \draw[-Latex, very thick] (O)--(axis cs:2,-3) node[right] {\footnotesize \(\begin{bmatrix}2 \\ -3\end{bmatrix}\)};
        \draw[red, -Latex, very thick] (O)--(axis cs:-4,6) node[left] {\footnotesize \(-2\begin{bmatrix}2 \\ -3\end{bmatrix}\)};
      \end{axis}
    \end{tikzpicture}

    \begin{tikzpicture}
      \begin{axis}[basic curve, black, xmin=-6, xmax=5, ymin=-8, ymax=0, xtick distance={2}, ytick distance={2}]
        \addplot coordinates {(0,0)};
        \coordinate (O) at (axis cs:0,0);
        \draw[-Latex, very thick] (O)--(axis cs:-3/2,-4) node[left] {\footnotesize \(\begin{bmatrix}-3/2 \\ -4\end{bmatrix}\)};
        \draw[-Latex, very thick] (O)--(axis cs:3,-1/2) node[below] {\footnotesize \(\begin{bmatrix}3 \\ -1/2\end{bmatrix}\)};
        \draw[-Latex, very thick, dashed] (axis cs:-3/2,-4) -- (axis cs:{-3/2+3},{-4-1/2});
        \draw[red, -Latex, very thick] (O)--(axis cs:{-3/2+3},{-4-1/2}) node[below] {\footnotesize the sum};
      \end{axis}
    \end{tikzpicture}
    \begin{tikzpicture}
      \begin{axis}[width=5in, basic curve, black, xmin=-5, xmax=5, ymin=-8, ymax=8, xtick distance={3}, ytick distance={3}]
        \addplot coordinates {(0,0)};
        \coordinate (O) at (axis cs:0,0);
        \draw[-Latex, very thick] (O)--(axis cs:1,-2) node[below right] {\footnotesize \(\begin{bmatrix}1 \\-2\end{bmatrix}\)};
        \draw[-Latex, very thick] (O)--(axis cs:2,3) node[above] {\footnotesize \(\begin{bmatrix}2 \\3\end{bmatrix}\)};
        \draw[-Latex, very thick, dotted] (O)--(axis cs:-2,-3) node[below left] {\footnotesize \(\begin{bmatrix}-2 \\-3\end{bmatrix}\)};
        \draw[-Latex, very thick, dashed] (axis cs:1,-2) -- (axis cs:{1-2},{-2-3});
        \draw[red, -Latex, very thick] (O)--(axis cs:{1-2},{-2-3}) node[below] {\footnotesize the difference};
      \end{axis}
    \end{tikzpicture}
  \end{solution}
\end{exercise}
\clearpage

\begin{exercise}
  Find \(a\) that satisfies the matrix-vector equation \(\begin{bmatrix} 1 & a \\ 0 & -1 \end{bmatrix} \begin{bmatrix} 3 \\ 7 \end{bmatrix} = \begin{bmatrix} 38 \\ -7 \end{bmatrix}\).

  \begin{solution}
    We perform the matrix multiplication to get a relation on \(a\). Note that
    \[
      \begin{bmatrix} 1 & a \\ 0 & -1 \end{bmatrix} \begin{bmatrix} 3 \\ 7 \end{bmatrix} = 
      \begin{bmatrix}
        3 + 7a \\
        0 - 7
      \end{bmatrix}.
    \]
    For this vector to be equal to \(\begin{bmatrix} 38 \\ -7\end{bmatrix}\), we must have \(3 + a \times 7 = 38\). Therefore, \(a = 5\).
  \end{solution}
\end{exercise}

\begin{exercise}
  Consider the following transition diagram that models an age-structured population where \(x_{1}(t), x_{2}(t), x_{3}(t)\) count juveniles, adolescence and adults at time \(t\), respectively.

  \begin{tikzpicture}[transition diagram]
    \node[class] (A) at ( 120:1) {\(x_{1}\)};
    \node[class] (B) at (   0:1) {\(x_{2}\)};
    \node[class] (C) at ( 240:1) {\(x_{3}\)};

    % forward loop
    \path[->] (A) edge [bend  left] node[above] {\(70\%\)} (B);
    \path[->] (B) edge [bend  left] node[below] {\(50\%\)} (C);
    \path[->] (C) edge [bend  left] node[ left] {\(80\%\)} (A);

    % backwards loop
    % \path[->] (A) edge [bend  left] node[ left] {AC} (C);
    % \path[->] (C) edge [bend  left] node[below] {CB} (B);
    \path[->] (B) edge [bend  left] node[above] {\(p\)} (A);

    \path[->] (A) edge [loop  left] node[left ] {\(10\%\)} (A);  
    \path[->] (B) edge [loop right] node[right] {\(20\%\)} (B);  
    % \path[->] (C) edge [loop  left] node[left ] {CC} (C);
  \end{tikzpicture}

  \begin{enumerate}
    \item Describe the transition from juveniles to adolescence in words. How many juveniles becomes adults in one time tick? How many juveniles do not survive in one time tick? How many juveniles stay as juveniles in one time tick?
    \item What portion of the juveniles do not survive to become adolescence?
    \item What does the arrow labelled \(p\) represent?
    \item What are the birth and death rates in adults?
    \item Translate the diagram into a (vector) recurrence.
    \item Describe the transition \emph{between} any two subpopulation in words.
    \item Assume \(p = 0\). Find \(x_{3}(2)\) if \(\vec{x}(0) = \begin{bmatrix} 0 \\ 100 \\ 200 \end{bmatrix}\).
  \end{enumerate}

  \begin{solution} \phantom{top}

    \begin{enumerate}
      \item \(70\%\) of the juveniles grow into adolescence. No juveniles grow into adults in one time increment. \(10\%\) of the juveniles stay as juveniles in one time increment.
      \item \(70\% + 10\%\) of the juveniles survive. Therefore, \(20\%\) of the juveniles do not survive.
      \item The arrow labelled \(p\) means the birth rate among adolescence.
      \item The birth rate among adults is \(80\%\). The death rate is unknown.
        \hlwarn{Correction: the death rate is determinable. Since adults have no self-loop (the \((3,3)\) entry of \(A\) is \(0\)), no adults survive, so the death rate is \(100\%\) --- as the next item states.}
      \item The birth rate among adults is \(80\%\). The death rate is \(100\%\).
      \item In addition to the above, \(20\%\) of the adolescence remain as adolescence, and \(50\%\) grow into adults.
      \item The model \(\vec{x}(t+1) = A\vec{x}\) has matrix 
        \[
          A = 
          \begin{bmatrix}
            0.1 &   p & 0.8 \\
            0.7 & 0.2 & 0   \\
            0   & 0.5 & 1   \\
          \end{bmatrix}
          =
          \begin{bmatrix}
            0.1 & 0   & 0.8 \\
            0.7 & 0.2 & 0   \\
            0   & 0.5 & 0   \\
          \end{bmatrix}.
        \]
        Note 
        \(
          \vec{x}(2) = 
          \begin{bmatrix}
            50 \\
            116 \\
            10
          \end{bmatrix}
        \). Therefore, \(x_{3}(2) = 10\).

        \hlwarn{Correction: the first entry of \(\vec{x}(2)\) should be \(56\), not \(50\) (since \(0.1 \cdot 160 + 0.8 \cdot 50 = 56\)). So \(\vec{x}(2) = [56, 116, 10]^{\top}\). The answer \(x_{3}(2) = 10\) is unaffected.}
    \end{enumerate}
  \end{solution}
\end{exercise}

\begin{exercise}
  Vector recurrence and transition diagrams can do more than modelling age-structure populations. The idea can be applied to model transitions among any states. 

  Let \(x_{1}(t), x_{2}(t), x_{3}(t)\) represent the probability of rain, snow and no precipitation at time \(t\) (measured in hours).  
  \begin{itemize}
    \item If it is raining at time \(t\), then the probability of rain, snow and no precipitation at time \(t+1\) is \(0.6, 0.1, 0.3\) respectively.
    \item If it is snowing at time \(t\), then the probability of rain, snow and no precipitation at time \(t+1\) is \(0.2, 0.5, 0.3\) respectively.
    \item If it is neither raining nor snowing at time \(t\), then the probability of rain, snow and no precipitation at time \(t+1\) is \(0.1, 0.1, 0.8\) respectively.
  \end{itemize}
  Assume further that these are the only three weather states possible.  
  \begin{itemize}
    \item Sketch the transition diagram for the three weather states.
    \item Model the \(\vec{x}(t)\) as a vector recurrence.  
    \item Find \(\vec{x}(2)\). What do its entries represent?
  \end{itemize}

  \begin{solution}
    The arrows in the middle point counterclockwise.

    \begin{tikzpicture}[transition diagram]
      \node[class] (A) at ( 120:1) {rain\\\(x_{1}\)};
      \node[class] (B) at (   0:1) {snow\\\(x_{2}\)};
      \node[class] (C) at ( 240:1) {no precip.\\\(x_{3}\)};

      \path[->] (A) edge [loop  left] node[left ] {\(60\%\)} (A);  
      \path[->] (A) edge [bend  left] node[above] {\(10\%\)} (B);
      \path[->, red] (A) edge [bend  left] node[ left] {\(30\%\)} (C);

      \path[->, blue] (B) edge [bend  left] node[above] {\(20\%\)} (A);
      \path[->] (B) edge [loop right] node[right] {\(50\%\)} (B);  
      \path[->] (B) edge [bend  left] node[below] {\(30\%\)} (C);

      \path[->] (C) edge [bend  left] node[left]  {\(10\%\)} (A);
      \path[->] (C) edge [bend  left] node[below] {\(10\%\)} (B);
      \path[->] (C) edge [loop  left] node[left ] {\(80\%\)} (C);  
    \end{tikzpicture}

    The recurrence is 
    \[
      \vec{x}(t+1) = 
      \begin{bmatrix}
        0.6 & 0.2 & 0.1 \\
        0.1 & 0.5 & 0.1 \\
        0.3 & 0.5 & 0.8 \\
      \end{bmatrix}
      \vec{x}(t).
    \]

    \hlwarn{Correction: the \((3,2)\) entry should be \(0.3\), not \(0.5\) (from snow, no-precipitation has probability \(0.3\); each column must sum to \(1\)).}
  \end{solution}

  We are not given numbers for \(\vec{x}(0)\), so let \(\vec{x}(0) = \begin{bmatrix} a \\ b \\ c \end{bmatrix}\) where \(a,b,c\) are constants (i.e., placeholders). Then
  \[
    \vec{x}(2) = A^{2} \vec{x}(0) = 
    \begin{bmatrix}
      0.41 & 0.27 & 0.09 \\
      0.14 & 0.32 & 0.07 \\
      0.26 & 0.36 & 0.09
    \end{bmatrix}
    \begin{bmatrix}
      a \\
      b \\
      c 
    \end{bmatrix} =
    \begin{bmatrix}
      0.41 a + 0.27 b + 0.09 c \\
      0.14 a + 0.32 b + 0.07 c \\
      0.26 a + 0.36 b + 0.09 c
    \end{bmatrix}.
  \]
  The entries of \(\vec{x}(2)\) represent the probabilities of rain, snow and no precipitation at time \(t = 2\).

  \hlwarn{Correction: using the corrected matrix, \(A^{2} = \begin{bmatrix} 0.41 & 0.25 & 0.16 \\ 0.14 & 0.30 & 0.14 \\ 0.45 & 0.45 & 0.70 \end{bmatrix}\) (each column sums to \(1\)); the displayed \(A^{2}\) is incorrect. Update \(\vec{x}(2) = A^{2}\vec{x}(0)\) accordingly.}
\end{exercise}

\begin{exercise}
  A company tracks its workforce in three career-stage categories.
  Suppose \(x_{1}(t), x_{2}(t), x_{3}(t)\) represent the number of entry-level, mid-career and senior employees at year \(t\).  
  \begin{itemize}
    \item Senior employees generate \(1.2\) new entry-level hires per senior employee per year.
    \item \(60\%\) of entry-level employees are promoted to mid-career roles each year. The rest leave the company.
    \item \(70\%\) of mid-career employees are promoted to a senior positions each year. \(10\%\) leave the company. The rest stay without promotions.
    \item \(90\%\) of senior employees stays with the business every year.
  \end{itemize}
  Initially, there are \(300\) entry-level, \(200\) mid-career and \(100\) senior employees.  Sketch the transition diagram and model the workforce as a vector recurrence.

  \begin{solution}
    The arrows in the middle point counterclockwise.

    \begin{tikzpicture}[transition diagram]
      \node[class] (A) at ( 120:1) {entry-level\\\(x_{1}\)};
      \node[class] (B) at (   0:1) {mid-career\\\(x_{2}\)};
      \node[class] (C) at ( 240:1) {senior\\\(x_{3}\)};

      % \path[->] (A) edge [loop  left] node[left ] {\(40\%\)} (A);  
      \path[->] (A) edge [bend  left] node[above] {\(60\%\)} (B);
      % \path[->, red] (A) edge [bend  left] node[ left] {\(30\%\)} (C);

      % \path[->, blue] (B) edge [bend  left] node[above] {\(20\%\)} (A);
      \path[->] (B) edge [loop right] node[right] {\(20\%\)} (B);  
      \path[->] (B) edge [bend  left] node[below] {\(70\%\)} (C);

      \path[->] (C) edge [bend  left] node[left]  {\(1.2\)} (A);
      % \path[->] (C) edge [bend  left] node[below] {\(10\%\)} (B);
      \path[->] (C) edge [loop  left] node[left ] {\(90\%\)} (C);  
    \end{tikzpicture}

    The model is 
    \[
      \vec{x}(t+1) = 
      \begin{bmatrix}
        0   & 0   & 1.2 \\
        0.6 & 0.2 & 0   \\
        0   & 0.7 & 0.9
      \end{bmatrix}
      \vec{x}(t),
      \quad\text{with}\quad
      \begin{bmatrix}
        300 \\
        200 \\
        100   
      \end{bmatrix}.
    \]
  \end{solution}
\end{exercise}

\begin{exercise}
  A biologist is studying the population dynamics of a species of frogs with three subpopulations tadpoles \(x_{1}(t)\), juveniles \(x_{2}(t)\) and adults \(x_{3}(t)\). Suppose the population dynamics are modelled by the recurrence
  \[
    \vec{x}(t+1) =
    \begin{bmatrix}
      0 & 0 & 120 \\
      0.25 & 0 & 0 \\
      0 & 0.4 & 0.85
    \end{bmatrix}
    \vec{x}(t).
  \]
  Describe, in words, the transition among subpopulations from year \(t\) to year \(t + 1\).

  \begin{solution}
    At the transition from year \(t\) to \(t+1\),
    \begin{itemize}
      \item each adult gives birth to \(120\) tadpoles, i.e., the birth rate among adult frogs is \(12000\%\),
      \item \(25\%\) of tadpoles grows into juveniles, 
      \item \(40\%\) of the juveniles grows into adults, and
      \item \(85\%\) of the adults remain as adults.
    \end{itemize}
  \end{solution}
\end{exercise}

\begin{exercise}
  Consider a vector recurrence \(\vec{x}(t+1) = A \vec{x}\).  The arrows in its vector field are described by \(\vec{v} = \begin{bmatrix} 1 & 1 \\ -1 & 1 \end{bmatrix}\vec{x}\) where \(\vec{x}\) represents arbitrary points in the plane. In other words, \(\vec{v}\) is an arrow in the vector field drawn by placing its tail at the tip of \(\vec{x}\). Find \(A\).

  \begin{solution}
    The arrows \(\vec{v}\) in the vector field represent changes from \(\vec{x}(t)\) to \(\vec{x}(t+1)\). An arbitrary state \(\vec{x}(t)\) is just any vector, denoted as \(\vec{x}\) Therefore, they are
    \[
      \vec{v} = \vec{x}(t+1) - \vec{x}(t) = A \vec{x}(t) - I \vec{x}(t) = (A - I) \vec{x}.
    \]
    Using the given relation between \(\vec{v}\) and \(\vec{x}\), we get relation
    \( A - I = \begin{bmatrix} 1 & 1 \\ -1 & 1 \end{bmatrix} \). Therefore, \(A = \begin{bmatrix} 2 & 1 \\ -1 & 2 \end{bmatrix}\).
  \end{solution}
\end{exercise}

\begin{exercise}
  Find the unknown constants \(a,b\) in the following equations.

  \begin{itemize}
    \item \(\begin{bmatrix} -13 \\ 18 \end{bmatrix} = a \begin{bmatrix} -2 \\ 5 \end{bmatrix} + b \begin{bmatrix} 3 \\ 7 \end{bmatrix}\).
      % a = 5, b = -1
    \item \(\begin{bmatrix} 1/2 \\ -7/3 \end{bmatrix} = a \begin{bmatrix} 2 \\ -1 \end{bmatrix} + b \begin{bmatrix} -3 \\ 4 \end{bmatrix}\).
      % a = 1/2, b = 2
    \item \(\begin{bmatrix} -5/6 \\ 7/15 \end{bmatrix} = a \begin{bmatrix} 1 \\ 1/5 \end{bmatrix} + b \begin{bmatrix} 1/2 \\ 2 \end{bmatrix}\).
      % a = -1, b = 1/3
    \item \(\begin{bmatrix} 4 \\ 9 \\ -7 \end{bmatrix} = a \begin{bmatrix} 1 \\ -3 \\ 2 \end{bmatrix} + b \begin{bmatrix} -2 \\ 5 \\ 1 \end{bmatrix} + c \begin{bmatrix} 0 \\ 4 \\ -3 \end{bmatrix}\).
  \end{itemize}

  \begin{solution}
    Solving the unknown constants amounts to solving a system of linear equations.

    \begin{itemize}
      \item \(a = 5, b = 1\).
        \hlwarn{Correction: \(b = -1\), not \(1\) (check: \(-2 \cdot 5 + 3 \cdot (-1) = -13\) and \(5 \cdot 5 + 7 \cdot (-1) = 18\)).}
        % sage: A = matrix([[-2,3],[5,7]])
        % sage: v = vector([-13,18])
        % sage: A.inverse()*v

      \item \(a = -1, b = -5/6\).
        % sage: A = matrix([[2,-3],[-1,4]])
        % sage: v = vector([1/2,-7/3])
        % sage: A.inverse()*v

      \item \(a = -1, b = 1/3\).
        % sage: A = matrix([[1,1/2],[1/5,2]])
        % sage: v = vector([-5/6,7/15])
        % sage: A*v

      \item \(a = 74/17, b = 3/17, c = 90/17\).
        % sage: A = matrix([[1,-2, 0],[-3,5,4], [2,1,-3]])
        % sage: v = vector([4,9,-7])
        % sage: A.inverse()*v
    \end{itemize}
  \end{solution}
\end{exercise}
\end{document}

% transition diagram template
\begin{tikzpicture}[transition diagram]
  \node[class] (A) at ( 120:1) {A};
  \node[class] (B) at (   0:1) {B};
  \node[class] (C) at ( 240:1) {C};

  % forward loop
  \path[->] (A) edge [bend  left] node[above] {AB} (B);
  \path[->] (B) edge [bend  left] node[below] {BC} (C);
  \path[->] (C) edge [bend  left] node[left]  {CA} (A);

  % forward loop
  \path[->] (A) edge [bend  left] node[ left] {CA} (C);
  \path[->] (C) edge [bend  left] node[below] {CB} (B);
  \path[->] (B) edge [bend  left] node[above] {BA} (A);

  \path[->] (A) edge [loop  left] node[left ] {AA} (A);  
  \path[->] (B) edge [loop right] node[right] {BB} (B);  
  \path[->] (C) edge [loop  left] node[left ] {CC} (C);  
\end{tikzpicture}
